%  ========================================================================
%  Copyright (c) 1985 The University of Washington
%
%  Licensed under the Apache License, Version 2.0 (the "License");
%  you may not use this file except in compliance with the License.
%  You may obtain a copy of the License at
%
%      http://www.apache.org/licenses/LICENSE-2.0
%
%  Unless required by applicable law or agreed to in writing, software
%  distributed under the License is distributed on an "AS IS" BASIS,
%  WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
%  See the License for the specific language governing permissions and
%  limitations under the License.
%  ========================================================================
%

% Documentation for University of Washington thesis LaTeX document class
% by Jim Fox
% fox@washington.edu
%
%    Revised 2020/02/24, added \caption()[]{} option.  No ToC.
%
%    Revised for version 2015/03/03 of uwthesis.cls
%    Revised, 2016/11/22, for cleanup of sample copyright and title pages
%
%    This document is contained in a single file ONLY because
%    I wanted to be able to distribute it easily.  A real thesis ought
%    to be contained on many files (e.g., one for each chapter, at least).
%
%    To help you identify the files and sections in this large file
%    I use the string '==========' to identify new files.
%
%    To help you ignore the unusual things I do with this sample document
%    I try to use the notation
%       
%    % --- sample stuff only -----
%    special stuff for my document, but you don't need it in your thesis
%    % --- end-of-sample-stuff ---


%    Printed in twoside style now that that's allowed
%
 
\documentclass [11pt, proquest] {uwthesis}[2020/02/24]
 
%
% The following line would print the thesis in a postscript font 

\usepackage{booktabs}     % Professional tables
\usepackage{graphicx}     % For figures
\usepackage{amsmath}      % Math equations
\usepackage{hyperref}     % Hyperlinks in PDF
\usepackage[colorinlistoftodos,prependcaption]{todonotes}  % Drafting notes
% Usage: \todo{Note text} for margin notes
%        \todo[inline]{Note text} for inline notes
%        \listoftodos to generate list of all todos

\setcounter{tocdepth}{1}  % Print the chapter and sections to the toc
 

% ==========   Local defs and mods
%

% --- sample stuff only -----
% These format the sample code in this document

\usepackage{alltt}  % 
\newenvironment{demo}
  {\begin{alltt}\leftskip3em
     \def\\{\ttfamily\char`\\}%
     \def\{{\ttfamily\char`\{}%
     \def\}{\ttfamily\char`\}}}
  {\end{alltt}}
 
% metafont font.  If logo not available, use the second form
%
% \font\mffont=logosl10 scaled\magstep1
\let\mffont=\sf
% --- end-of-sample-stuff ---
 



\begin{document}
 
% ==========   Preliminary pages
%
% ( revised 2012 for electronic submission )
%

\prelimpages
 
%
% ----- copyright and title pages
%
\Title{Cross-Detector Descriptor Fusion: \\
Scale Control and Spatial Alignment for Local Feature Matching}
\Author{Frank Sossi}
\Year{2026}
\Program{Computer Science \& Software Engineering}

\Chair{Professor Clark Olson}{Committee Chair}{Computing \& Software Systems}
\Signature{Professor Min Chen}
\Signature{Professor Dong Si}

\copyrightpage

\titlepage  

 
%
% ----- signature and quoteslip are gone
%

%
% ----- abstract
%


\setcounter{page}{-1}
\abstract{%
% TODO: Check abstract length against UW requirements (currently ~340 words, typical limit 150-350)
% TODO: Verify all numbers match Results chapter
Local feature descriptors are fundamental to many computer vision applications including SLAM, structure-from-motion, and image retrieval. This thesis investigates two complementary approaches to improving local feature matching: using multiple detectors as a quality filter for keypoint selection, and fusing complementary descriptors to combine their strengths.

We present four main contributions. First, we demonstrate that spatial intersection between distinct keypoint detectors acts as a powerful quality filter. When different detection methods---whether SIFT and SURF or SIFT and KeyNet---both identify a keypoint at the same location, this consensus indicates a genuinely distinctive feature. Descriptors computed at intersection keypoints consistently outperform those on single-detector sets, with HardNet achieving 82.4\% mAP on SIFT-KeyNet intersection---a 25\% relative improvement and the best single-descriptor result in our study.

Second, we create a color version of the HPatches patch benchmark, enabling evaluation of color-aware descriptors. Using this dataset, we show that fusing the color histogram descriptor HoNC with learned CNN descriptors yields substantial improvements: HoNC+SOSNet concatenation achieves 50.6\% mAP on patch matching, outperforming all individual descriptors. HoNC's strong discriminative capability (high verification-to-matching ratio) complements the CNN's matching-optimized representations.

Third, we show that cross-family fusion (SIFT+CNN) requires pre-fusion L2 normalization to ensure equal contribution from each descriptor; with proper normalization, SIFT+HardNet achieves 46.0\% mAP on patches. Fourth, we show that keypoint scale is a dominant factor: filtering to large-scale keypoints yields 39\% relative improvement for SIFT and 21\% for CNN descriptors.

We develop DescriptorWorkbench, an open-source evaluation framework, and conduct over 100 experiments. The results demonstrate that keypoint quality---determined by detector consensus and scale---has greater impact on matching performance than descriptor algorithm choice alone.
}
 
%
% ----- contents & etc.
%
\tableofcontents
\listoffigures
\listoftables

% TODO: Remove \listoftodos before final submission
\listoftodos
 
%
% ----- acknowledgments
%
\acknowledgments{% \vskip2pc
  % {\narrower\noindent
  I would like to thank Professor Clark Olson for his guidance and support throughout this research.
  % \par}
}

%
% ----- dedication
%
% \dedication{\begin{center}to my dear wife, Joanna\end{center}}

%
% end of the preliminary pages
 
 
 
%
% ==========      Text pages
%

\textpages
 
% ========== Chapter 1
\chapter{Introduction}
\label{chap:introduction}
\input{sections/introduction}

% ========== Chapter 2
\chapter{Background and Related Work}
\label{chap:background}
\input{sections/background}

% ========== Chapter 3
\chapter{Methodology}
\label{chap:methodology}
\input{sections/methodology_new}

% ========== Chapter 4
\input{sections/results}

% ========== Chapter 5
\chapter{Discussion}
\label{chap:discussion}
\input{sections/discussion}

% ========== Chapter 6
\chapter{Conclusion and Future Work}
\label{chap:conclusion}
\input{sections/conclusion}

%
% ==========   Bibliography
%
\nocite{*}   % include everything in the uwthesis.bib file
\bibliographystyle{plain}
\bibliography{uwthesis}

\end{document}
